Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-17 Manuscript/manuscript.Rmd
manuscript and abstract edits
Trying to read

To investigate the relationship between functional genomic annotations and the development of TAD boundaries, we devised 3 different predictor-types that composed the feature space of downstream random forest classification models. These predictor-type specific annotations represented the data matrix, **X** = {$X_{1}$, $X_{2}$, ..., $X_{p}$}, where *p* is the complete set of functional annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C).  When measured in basepairs, distances were found to be highly skewed across all resolutions, for both the GM12878 and K562 cell lines (Supplementary Figures 1-2 A-D). As a normalization procedure, we performed a $log_{2}$ transformation, which resulted in much more gaussian-shaped distance densities (Supplementary Figures 1-2 E-H).

@@

We then developed an ensemble framework to systematically compare and establish the best set of data level characteristics for predicting TAD boundaries using random forest classifiers built on cell line specific functional genomic annotations (Figure 3). TAD boundaries were called using the arrowhead algorithm [@durand2016juicebox] from replicate Hi-C data for 2 cell lines *L* = {*GM12878*, *K562*}, at four different resolutions *R* = {*10 kb*, *25 kb*, *50 kb*, *100 kb*}. For each classifier, we considered 3 separate predictor types as the feature space including *P* = {*OC*, *OP*, *Distance*}. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including *B* = {*None*, *RUS*, *ROS*, *SMOTE*}. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome, and thus, was repeated repeated 21 times (number of autosomal chromosomes, excluding chromosome 9), *C* = {*1*, *2*, ..., *22*}. A system produced from the ensemble framework was defined as: *S* = {*c*, *r*, *p*, *b*}, where *l* $\in$ *L*, *r* $\in$ *R*, *p* $\in$ *P*, *b* $\in$ *B*, and *c* $\in$ *C*. Letting *|X|* denote the cardinality of each set, we evaluated *|L|* $\times$ *|R|* $\times$ *|P|* $\times$ *|B|* different ensemble systems, totaling 2016 (= 2 $\times$ 4 $\times$ 3 $\times$ 4 $\times$ 21) separate models. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9; see Methods). We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.

@@

[???probably better to report medians here]
For GM12878, the average number of genomic bins created across chromosomes was 12,214.1, 4,892.7, 2,450.9, and 1,227.5 for each resolution respectively (Table 1). Likewise, the average number chromosome specific TADs called on each resolution was found to be 375.0, 228.4, 115.1, and 50.6. These TADs resulted in an average of 1254.6, 714.0, 364.4, and 165.3 chromosome specific boundaries across each resolution. Each of these quantities was find to have a significantly large ICC, indicating that the number of genomic bins, TADs called, and corresponding unique boundaries were highly correlated between the different resolutions (Table 2). On the other hand, we found that the chromosome specific imbalance ratios remained consistenly severe across all resolution, with averages of 0.12, 0.17, 0.17, and 0.15 for each resolution respectively (ICC=0.114; p=0.109). Such low imbalance ratios indicated a wide disparity between the number of genomic bins that overlaped with a flanked TAD boundary and the much larger number that did not. The K562 cell line was found to exhibit similar profiles for the above quantities (Supplementary Tables 3-4)

@@

Using the jaccard index, we were able to determine which group-specific annotations were enriched together on the linear genome. As could be expected, for GM12878, the class of 14 annotations segmenting the chromation into different states exhibited very low pairwise correlations, with a jaccard index range of (0-0.001) (Table 1). However, specific sets of pairwise histone modifications and transcription factor binding sites (TFBS) were found to be highly enriched. The range of jaccard indices for the 10 histone modifications was found to be (0.022-0.539). Specifically, H3k27ac, H3k4me1, H3k4me2, and H3k4me3 were found to be highly correlated, high levels of which have been detected surrounding transcription start sites [@barski2007high]. For the 79 TFBS, the range of jaccard indices was (<0.001-0.425). Here, known mediators of loop extrusion components were found to have high pairwise correlations including CTCF, RAD21, and SMC3 [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Similar profiles of jaccard indices were exhibited across the sets of annotations for the K562 cell line (Supplementary Table 4). <!--The high jaccard indices among select sets of genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation.-->

@@

We then went on the evaluate the effect of class imbalance on model performance. In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 5A). We gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Here, the results were similar to that of OC types (Figure 5B). However, when including distance-type predictors, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 5C). Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy, sensitivity and specificity, and AUC) were less sensitive for models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7).

@@

>>>>

To investigate the relationship between functional genomic annotations and the development of TAD boundaries, we devised 3 different predictor-types that composed the feature space of downstream random forest classification models. These predictor-type specific annotations represented the data matrix, **X** = {$X_{1}$, $X_{2}$, ..., $X_{p}$}, where $p$ is the complete set of functional annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C).  When measured in basepairs, distances were found to be highly skewed across all resolutions, for both the GM12878 and K562 cell lines (Supplementary Figures 1-2 A-D). As a normalization procedure, we performed a $log_{2}$ transformation, which resulted in much more gaussian-shaped distance densities (Supplementary Figures 1-2 E-H).

@@

We then developed an ensemble framework to systematically compare and establish the best set of data level characteristics for predicting TAD boundaries using random forest classifiers built on cell line specific functional genomic annotations (Figure 3). TAD boundaries were called using the Arrowhead algorithm [@durand2016juicebox] from replicate Hi-C data for 2 cell lines $L = \{GM12878, K562\}$, at four different resolutions $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. For each classifier, we considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. To assess the severity of class imbalance, four re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome, and thus, was repeated repeated 21 times (number of autosomal chromosomes, excluding chromosome 9 [??? In Methods, not here, you should explain why]), $C = \{1, 2, ..., 22\}$. A system produced from the ensemble framework was defined as: $S = \{c, r, p, b\}$, where $l \in L$, $r \in R$, $p \in P$, $b \in B$, and $c \in C$. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9; see Methods). We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials. [??? Currently, Methods and the results are redundant. Look how others present their models in the results section, and how methods section is structured in such cases. Specifically, some key definitions, like OC, OP, Distance should be solidly introduced in the results section, while log2 status of distance and supplementary figure 1 description are for methods]

@@

[???probably better to report medians here]
For GM12878, the average number of genomic bins created across chromosomes was 12,214.1, 4,892.7, 2,450.9, and 1,227.5 for each resolution respectively (Table 1). Likewise, the average number chromosome specific TADs called on each resolution was found to be 375.0, 228.4, 115.1, and 50.6. These TADs resulted in an average of 1254.6, 714.0, 364.4, and 165.3 chromosome specific boundaries across each resolution. Each of these quantities was find to have a significantly large ICC, indicating that the number of genomic bins, TADs called, and corresponding unique boundaries were highly correlated between the different resolutions (Table 2). On the other hand, we found that the chromosome specific imbalance ratios remained consistenly severe across all resolution, with averages of 0.12, 0.17, 0.17, and 0.15 for each resolution respectively (ICC=0.114; p=0.109). Such low imbalance ratios indicated a wide disparity between the number of genomic bins that overlaped with a flanked TAD boundary and the much larger number that did not. The K562 cell line was found to exhibit similar profiles for the above quantities (Supplementary Tables 3-4) [??? Very bad and boring. You should tell the STORY, not the report. E.g., "Expectedly, the average number of genomic bins, TADs, and TAD boundaries increased with the increased resolution of Hi-C data (Table 1). However, the imbalance ratio, defined as ..., remained severe (~XXX) at all resolutions (Table 1). In summary, the problem of class imbalance exists at all resolutions of Hi-C data."]

@@

Using the Jaccard [??? Spell capital everywhere] index, we were able to determine which group-specific annotations were enriched together on the linear genome. As could be expected, for GM12878, the class of 14 annotations segmenting the chromation into different states exhibited very low pairwise correlations, with a jaccard index range of (0-0.001) (Table 1). [??? Wrong table?] However, specific sets of pairwise histone modifications and transcription factor binding sites (TFBS) were found to be highly enriched. The range of jaccard indices for the 10 histone modifications was found to be (0.022-0.539). Specifically, H3k27ac, H3k4me1, H3k4me2, and H3k4me3 were found to be highly correlated, high levels of which have been detected surrounding transcription start sites [@barski2007high]. For the 79 TFBS, the range of jaccard indices was (<0.001-0.425). Here, known mediators of loop extrusion components were found to have high pairwise correlations including CTCF, RAD21, and SMC3 [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Similar profiles of jaccard indices were exhibited across the sets of annotations for the K562 cell line (Supplementary Table 4). <!--The high jaccard indices among select sets of genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation.-->

@@

We then went on the evaluate the effect of class imbalance on model performance. In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 5A). We gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Here, the results were similar to that of OC types (Figure 5B). However, when including distance-type predictors, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 5C). Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy, sensitivity and specificity, and AUC) were less sensitive for models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7). [??? Need to know why PRCs give different results.]

@@

After determining the optimal set of data level characteristics necessary for predicting TAD boundaries, we then aimed to identify the set of functional genomic elements that were most influential in predicting said boundaries.
